---
title: "Base R => Tidy R"
output:
  html_document:
    toc: true
    toc_float: true
    # df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Dependencies
```{r deps}
library(tidyverse)
```

**Note**: For all snippets, df is set to the mtcars dataset.

### Basics
<!-- {.tabset} -->

#### Data frame vs tibble
**Printing**
```{r print}
# Base R prints all rows
mtcars
# Tidy R does not
tbl_df(mtcars)
# If you want custom print
print(as.tibble(mtcars), n=11)
```
**Convert one to the other**
```{r conversion}
df <- mtcars
# convert df to tbl
as.tibble(df) %>% head()
# back to df
as.tibble(df) %>% as.data.frame() %>% head()
```
**Data frame converts characters to factors, tibble doesn't ever convert type**
```{r conversion_types}
# data frame converts char to factors
df <- data.frame(
  `:)` = "smile", 
  ` ` = "space",
  `2000` = "number")
str(df)
# tibble doesn't do this
tb <- tibble(
  `:)` = "smile", 
  ` ` = "space",
  `2000` = "number"
)
str(tb)
```

**tidyverse tip**: When dealing with the tidyverse functions, the grammar is very consistent with the verbs. The first argument always takes in the data followed by the arguments of the verb. So when using the function as is (w.o the piping operator `%>%`), supply data frame as first argument. For eg: `slice(df, 1:5)` versus `df %>% slice(1:5)`

#### Vector
**Grab vector from a subset**
```{r grab_vec}
df <- mtcars
# Base R
df[1:5, 'mpg']
# Tidy R
df %>% slice(1:5) %>% pull(mpg)
# or less idiomatic
# The dot stands as a placeholder for the data being piped in by `%>%` operator
# For e.g. df %>% .[['mpg']] is equivalent to df[['mpg']]
df %>% slice(1:5) %>% .$mpg
```

#### Slice 
```{r slice}
df <- mtcars
# Base R
df[1:5, ]
# Tidy R
df %>% slice(1:5)
```

**Q**: Why are the row names missing? (version 1)

**A**: All data frames have a row names attribute, a character vector of length the number of rows with no duplicates nor missing values. **Gotcha**: The `slice` verb silently drops the row names column.

**Fix**: Use `rownames_to_columns()` to explicitly save rownames as a column and `column_to_rownames()` to convert the column back to the row names attribute.

```{r rownames}
df <- mtcars
df %>% 
	rownames_to_column() %>%
	slice(1:5) %>%
	column_to_rownames()
```

**Q**: Why are the row names missing? (version 2)

**A**: `column_to_rownames()` was called on a `tibble` which is the data type output of `slice`. Remember that tibbles do not have a special row name attribute so it removes it.

**Gotcha**: Calling `column_to_rownames()` on a tibble will not restore row names of original data frame. Make sure to call `column_to_rownames()` on data frame.
 
**Fix**: Use `as.data.frame()` after `slice` to convert a tibble to a data frame before calling `column_to_rownames()`.
```{r rownames_df}
df %>% 
	rownames_to_column() %>%
	slice(1:5) %>%
  as.data.frame() %>%
	column_to_rownames()
```
**Note:** Version 1 and 2 can be combined by giving a more complete answer of anticipating the mistake in version 1 (calling `column_to_rownames()` on tibble). However, they're written out as two different versions so it's clear that one still runs into trouble if even if calling the `rownames_to_column` and `column_to_rownames()` functions; an extra function was needed to resolve this.

### Select
<!-- {.tabset} -->

#### Keep columns based on list of labels
```{r select}
# Base R
df <- mtcars
df <- df[, c('mpg', 'cyl', 'wt')]
head(df)
# Tidy R w.o %>%
df <- mtcars
# Tidy R w %>%
# head() for less printing
select(df, c('mpg', 'cyl', 'wt')) %>% head()
df %>% select(c('mpg', 'cyl', 'wt')) %>% head()
```

#### Possible Errors
```{r select_errors, error=TRUE}
# not using %>% or missing data frame as first argument
df <- mtcars
select(mpg, cyl, wt)
df <- mtcars
select(c('mpg', 'cyl', 'wt'))
# w. %>%
# supplying df for verbs while using %>% syntax
df <- mtcars %>%
	select(df, mpg, cyl, wt)
df <- mtcars %>%
	select(df, c('mpg', 'cyl', 'wt'))
```

#### Drop columns based on list of labels
```{r drop_cols}
# Base R
df <- mtcars
df <- df[, !(names(df) %in% c('mpg', 'cyl', 'wt'))]
head(df)
# Tidy R
df <- mtcars %>% 
	select(-mpg, -cyl, -wt)
head(df)
# or using one of the helper function of select: `one_of`
df <- mtcars %>% 
	select(-one_of(c('mpg', 'cyl', 'wt')))
head(df)
# Equivalent
df <- mtcars %>% 
	select(-one_of('mpg', 'cyl', 'wt'))
head(df)
# Gotcha: must supply a char vector or list of quoted items, not a list of unquoted col vars
# df <- mtcars %>% 
# 	select(-one_of(mpg, cyl, wt))
# Other helper functions for select: https://dplyr.tidyverse.org/reference/select_helpers.html
# Error: All arguments must be character vectors, not integer
# General confusion might be what verbs and helper functions accept lists of vars or char vectors?
# see confusion here: https://stackoverflow.com/questions/35839408/r-dplyr-drop-multiple-columns
```

### Filter 
<!-- {.tabset} -->

#### Filter rows based on single condition
```{r filter}
# Base R
df <- mtcars
df <- df[df$cyl == 6, ]
head(df)
# Tidy R
df <- mtcars %>% 
	filter(cyl == 6)
head(df)
```

#### Filter rows based on multiple conditions
```{r filter_multiple}
# Base R
df <- mtcars
df <- df[df$cyl == 6 & df$mpg >= 20, ]
head(df)
# Tidy R
df <- mtcars %>% 
	filter(cyl == 6 & mpg >= 20)
head(df)
# & -> comma
df <- mtcars %>% 
	filter(cyl == 6, mpg >= 20)
head(df)
```

#### Filter using `between` helper function
```{r filter_between}
# Being more succinct, matter of style but can be nice recommendation
# Filter using condition of form: x >= left & x <= right
df <- mtcars
df <- df[df$hp >= 200 & df$hp <= 300, ]
head(df)
# use a filter helper function: `between`
df <- mtcars %>% 
	filter(between(hp, 200, 300))
head(df)
```

### Arrange
```{r arrange}
# Desc for mpg, if tie asc for wt (weight)
df <- mtcars
df <- df[order(-df$mpg, df$wt), ]
head(df)
df <- mtcars %>%
	arrange(-mpg, wt)
head(df)
# or use desc for - which is doing the same thing for e.g. desc(1:2) -> -1, -2
df <- mtcars %>%
	arrange(desc(mpg), wt)
head(df)
```

